\chapter{Memory and Allocation Safety}

\section{Caller-Owned Items}

heapx never owns the objects stored in the heap.
The library owns only backend metadata: array entries, locators, node wrappers, handle slots, and temporary tables.
This distinction is essential for a generic C library because the library cannot know how caller objects should be copied or destroyed.

Destroying a heap releases heap-owned metadata.
It does not call destructors for stored item pointers.

\section{Overflow-Aware Size Arithmetic}

Allocation sizes are checked centrally through two helpers:

\begin{lstlisting}[style=cstyle, caption={Centralized size arithmetic helpers.}]
int heapx_size_mul(size_t left, size_t right, size_t *out);
int heapx_size_add(size_t left, size_t right, size_t *out);
\end{lstlisting}

These helpers return \texttt{-1} on overflow.
They are used in library allocation paths, tests, and benchmarks.
This matters because the project intentionally supports scalable experiments: heap item counts, DIMACS graph sizes, CSR arrays, handle tables, consolidation tables, and benchmark item arrays all require size arithmetic.

\section{Node Pool}

Pointer-heavy backends allocate many fixed-size node wrappers.
heapx uses a fixed-size node pool to recycle these objects.
The pool allocates blocks of object slots and maintains a free list.

\begin{lstlisting}[style=cstyle, caption={Node pool state.}]
struct heapx_node_pool {
    size_t object_size;
    size_t block_capacity;
    void *free_list;
    union heapx_pool_block *blocks;
};
\end{lstlisting}

The pool stores backend metadata only.
It does not own or inspect caller items.

\section{Alignment}

The pool rounds object sizes up to an alignment at least as strict as \texttt{void *} and \texttt{long double}.
The allocation block is represented by a union that forces suitable alignment before the object payload.

This is a low-level but important C detail.
Without it, pooled objects could be returned at addresses that are invalid for the backend structs stored there.

\section{Allocation Failure Policy}

heapx follows a conservative failure policy:

\begin{itemize}
    \item constructors return \texttt{NULL};
    \item insert operations return \texttt{-1};
    \item failed insertions do not transfer item ownership;
    \item handle attachment failures roll back backend metadata;
    \item temporary consolidation allocation failures preserve heap correctness.
\end{itemize}

The Fibonacci and Kaplan backends deserve special mention.
During consolidation, temporary arrays or reusable tables may need to grow.
If that allocation fails, the implementation chooses a correctness-preserving fallback.
The heap remains valid, but the ideal consolidation work may be reduced or postponed.

\section{What Is Not Provided}

heapx currently does not expose a custom allocator interface.
All heap-owned metadata uses the C allocation functions directly.
This keeps the API small, but it is a limitation if the library is later used in environments with strict allocation policies.

The project also does not shrink arrays or tables aggressively.
This is intentional for now: the current focus is correctness, stable behavior, and meaningful benchmark comparison rather than minimizing retained metadata after transient peaks.
